Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-18 doc/mc-paper.tex
Minor edits of appendix sections

%% Our experiments are designed to evaluate the following:
%% \begin{itemize}
%% \item
%%   whether or toolchain can handle challenging benchmark problems,
%% \item  
%% the effectiveness of some design choices and optimizations,
%% \item
%% the tradeoffs between the unverified and verified checker, and
%% \item
%% how the performance of our toolchain compares to that of the MICE model counter verifier.
%% \end{itemize}

@@

We step through a detailed example to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as the hint.  These resolve to the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.

@@

The two optimizations we have implemented exploit the power of our
general resolution framework to define new structures within the
proof. In particular, they create new product nodes that are not part
of the POG representation.

@@

The primary purpose of our experiments is to assess the ability of our
implementation of the toolchain illustrated in Figure~\ref{fig:chain}
to check the results from a possible knowledge compiler competition,
along the lines of the annual competitions in SAT solvers, model
counters, and other automated reasoning tools.  As secondary goals, we
seek insights into how possible algorithmic choices affect the
performance of the toolchain and how our toolchain compares to the
MICE framework for verifying model counters~\cite{fichte:sat:2022}.

@@

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generate d-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
d-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in pog $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we attempted
to use our proof generator and unverified checker, limiting proof
generation to 10,000 seconds.  We also attempted to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

Figure~\ref{fig:d4:crat} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for the
proof generator and checker (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144.
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster
with the generator and checker.  These were for problems having small numbers of
models (relative to the number of variables), and so most of the time
spent by both \dfour{} and our proof generator was in running a CDCL
solver to search the very sparse solution space.
\cadical{} generally outperforms the miniSAT-based solver used by
\dfour{}.  At the other extreme, one problem that required only 19
seconds for \dfour{} required 8657 seconds to generate a proof and 45
seconds to check it.  It is unclear why the proof generation required
so much time.  Overall the ratios between the combined generation plus
checking times versus the time for \dfour{} had a harmonic mean of
5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs use for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the
other 14 timed out.  Of the 15 that did not overflow the clause
counter, we were able to complete a one-sided verification for 9, but
the other 6 timed out.

@@

These experiments suggest that a hybrid of the two approaches might
yield the best results in terms of consistency, runtime, and proof
size.  It would begin the recursive descent of $\validate$ according
to a structural approach, but monolithically generate forward
implication proofs once it encounters sufficiently small subgraphs.

>>>>

@@

We step through portions of the file to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.

@@

The two optimizations we have implemented exploit the power of our
general resolution framework to define new structures within the
proof.
They create new product nodes that are not part
of the POG representation.

@@

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
  whether or toolchain can handle challenging benchmark problems,
\item  
the effectiveness of some design choices and optimizations,
\item
the tradeoffs between the unverified and verified checker, and
\item
how the performance of our toolchain compares to that of the MICE model counter verifier.
\end{itemize}

@@

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generate d-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
d-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in POG $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
our proof generator and unverified checker, limiting proof
generation to 10,000 seconds.  We ran the programs to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

Figure~\ref{fig:d4:crat} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for the
proof generator and checker (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster with the
generator and checker.  These were for problems having small numbers
of models (relative to the number of variables), and so most of the
time spent by both \dfour{} and our proof generator was in running a
CDCL solver to search the very sparse solution space.  \cadical{}
generally outperforms the miniSAT-based solver used by \dfour{}.  At
the other extreme, one problem that required only 19 seconds for
\dfour{} required 8657 seconds to generate a proof and 45 seconds to
check it.  This one particular benchmark appears to expose weak performance
limitations by our implementation of BCP\@.  Overall the ratios
between the combined generation plus checking times versus the time
for \dfour{} had a harmonic mean of 5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs use for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the other 14
timed out.  Of the 15 that did not overflow the clause counter, we
were able to complete a one-sided verification for 9, but the other 6
timed out.  Overall, we were able to provide some level of
verification for 94\% of the benchmark problems.

@@

These experiments suggest that a hybrid of the two approaches might
yield the best results in terms of consistency, runtime, and proof
size.  It would begin the recursive descent of $\validate$ according
to a structural approach, but monolithically generate
the proof once it encounters a sufficiently small subgraph.
